---
title: "Exercise 1"
---

```{r global_options, include=FALSE}
library(knitr)
library(ggplot2)
opts_chunk$set(fig.align="center", fig.height=4, fig.width=5.5)
```

*Enter your name here*

**This homework is due by Monday evening at 11:59pm. Please upload your .Rmd fild and your knited file to https://claremontmckenna.app.box.com/f/140969a32dea44e9ab6c0c00146d66e3.**


**Problem 1: (4 pts)** Demonstrate basic command of Markdown by creating a bulleted list with three items, a numbered list with three items, and a sentence that has one word in bold and one word in italics.

*Your text goes here.*

**Problem 2: (3 pts)** The `economics` dataset contains various time series data from the US economy:

```{r}
economics
```

Use ggplot to make a line plot of the total population (`pop`, in thousands) versus time (column `date`).

```{r}
# your code goes here
```

**Problem 3: (3 pts)** Again using the `economics` dataset, now make a scatter plot (using `geom_point()`) of the number of unemployed versus the total population (`pop`), and color points by date.

```{r}
# your code goes here
```
