---
title: "temp"
description: "General orientation for PSYC 167"
#author: ""
#date: ""
#order: 1 
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
#source(here::here("R", "_common.R"))
# from source
slide_order <- c("02-content.qmd", "01-content.qmd", "introduction.qmd")
slide_name <- basename(this.path::this.path())
slide_name <- "introduction.qmd"
slide_num <- match(slide_name, slide_order)
```

---
title: "Introduction"
order: 2
---

testing




Welcome to the orientation learning module for PSYC 167. We have a lot of ground to cover this semester and in return you will have much to showcase for yourself. 

If you are willing to embrace challenge and troubleshooting, you will walk away with:

- conceptual knowledge about data visualization processes and techniques;
- practical skills coding in R (viz., ggplot)
- hands-on experience working with an organization liaison (e.g., like working with a client);
- a public blog (optional) for an online profile and an active GitHub account and evidence of productivity


*Warning*: If you are not up for the challenge, you may feel overwhelmed, likely be miserable, and will want to give up. If that's you, there are other courses that may be a better fit for you. 

We will cover everything here and more on the first day of class. See you there.

## Goals

-   General course orientation
-   get you ready for the first week

## Read

Read the [syllabus for this course](https://www.slicesofdata/psyc167/syllabus.html).

## Getting started goals

There are many possibilities for technical difficulties as we get started. The computer room is equipped with the software that we need, however, I strongly encourage you to install the necessary software on your own computer.

Throughout the semester we will be using software called `R`, `Rstudio`, `quarto`, and `github`. All of this software is free and open-source, and can be installed on windows, mac, or linux.

## Week 1 assignment

1.  enter here

## More Goals for Week 1
