---
title: "Class Exercise: Read, Manipulate, and Summarize Data"
output:
  html_document:
    always_allow_html: yes
    toc: yes
    number_sections: yes
    code_folding: hide
  pdf_document:
    toc: yes
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



*Enter your name here*   


# **Load relevant libraries**

Load **{readr}**, **{magrittr}**, **{dplyr}**, and **{lubridate}**.

```{r}
library(magrittr)
library(dplyr)
library(readr)
library(lubridate)
```


# **Read a `.csv` file from a URL**

The following data are taken from [tf.tfrrs.org](https://tf.tfrrs.org/all_performances/CA_college_m_Claremont_Mudd_Scripps.html?list_hnd=600&season_hnd=131).

The file referenced at the url below contains data from a particular year and for men on the CMS track & field team who participated in an event during different meets. 

url: https://github.com/slicesofdata/dataviz23/raw/main/data/tfrrs/out_2010_CMS_100.csv

Read the this file and assign it to an object holding the data frame.

```{r}
file_name <- "out_2010_CMS_100.csv"
base_url <- "https://github.com/slicesofdata/dataviz23/raw/main/data/tfrrs/"
DAT <- readr::read_csv(paste(base_url, file_name, sep = ""))
```

# **Take a glympse at the structure of the data frame**

```{r}
dplyr::glimpse(DAT)
```


What are the names and types variables in the data frame?

Consider whether you understand the variables and what the values represent. If not, make a note because you may wish to ask the CMS T&F liaison these questions during a meeting. 



# **Mutating Variables**

Create a new variable to represent the year those meets took place (see file name) and another variable that represents the event type (see file name). You may also wish to create a variable for whether the athlete is either a Stag or Athena.

```{r}
DAT <- DAT %>%
  mutate(.,
         Year_Meet = 2010,
         Event = "100m",
         Team = "Stag"
         )
```


# **Summarize Data**

Summarize the data in at least two ways.

```{r}
DAT %>%
  summarise(.,
            Seconds = mean(Time)
            )
```


```{r}
DAT %>%
  group_by(., Athlete, Event) %>%
  summarise(.,
            Mean = mean(Time),
            Range = (max(Time) - min(Time)),
            #R = range(Time, na.rm = T)
            )


DAT %>%
  group_by(., Athlete, Event) %>%
  reframe(.,
            Mean = mean(Time),
            #Range = max(Time) - min(Time),
            Range = range(Time, na.rm = T)
            )

```

Did **{dplyr}** throw an error for the `summarize()` function? 



# **Summarize your finding**

Describe something that you found based on your summary.





# **Problem x: Write the modified data file**

Write the new file to your local disk, specifically at `/data` using the **{here}** library.

Do not save it as the same name.

```{r}

```




**Please knit an `html` file.** 
