---
title: "R Day"
output: html_document
---

In RMarkdown files (*.Rmd), you can add headers using the pound symbol (#). For example:

# This is my header. 

## And a smaller header.

Underneath headers, I can describe other things that I am doing.... In R script files (.R), however, # serves to comment out R code that you do not want to execute. Similarly, if you embed your R code in your RMarkdown file, the # will work in the same way as it would in an R script file. For example, I can insert a block of R code and show this here:

```{r}
# I use the pound symbol in R code to indicate that I don't want this line of code to run
# 4 + 4

# If I don't use the pound symbol, RMarkdown will run/execute the code.
8 + 8
```

If I need a certain library to execute code, I can load them. 
```{r}
library(psych)
library(moments) 
```

I can check to see what my working directory is to make sure it's correct.
```{r}
getwd() # This should be Psyc109 on your desktop
```

Because my data file is in my working directory, I can read in the csv file and assign the file to an R data frame objec with a certain name. I can read the data from the Center for Disease Control and assign that data frame to an object named cdc and make sure that it's a data frame.
```{r}
cdc <- read.csv("cdc.csv")
is.data.frame(cdc) # Determine if the data is a data frame object. Should return TRUE
```

I can look a the structure of the data frame so that I can see how many vector/variables there are and what type they are. And I can also take a look at what the data frame looks like by looking at the first or last several cases. 
```{r}
str(cdc) # Let's you know the type of the data (e.g., numeric, integer, Factor, etc.)
head(cdc) # Shows the first few cases of your data set
tail(cdc) # Shows the last few cases of your data set
dim(cdc) # Shows the dimensions of the data frame; 20,000 observations; 9 columns
names(cdc) # What are the column/variable names in the data frame?
```
I can also View the data frame in a more user-friendly viewer as well as examing the descriptive statistics of certain variables by specifying the data frame object (e.g., cdc) and the vector/variable I wish to examine.
```{r}
View(cdc) # View the data file
mean(cdc$weight) #Shows the Mean of the variable weight
median(cdc$weight) #Shows the Median of the variable weight
var(cdc$weight) #Shows the variance of the variable weight
sd(cdc$weight) #Shows the Standard Deviation of the variable weight
summary(cdc$weight) #Provides a summary of the variable weight
skewness(cdc$weight) #Shows the Skewness of the variable weight
kurtosis(cdc$weight) #Shows the Kurtosis of the variable weight

hist(cdc$weight) #Produces a simple histogram of the variable weight
```

Once I am done with my coding, I can save my RMarkdown file (*.Rmd) and click the Knit HTML button to produce a pretty HTML file. This file will also be saved in my working directory (Psyc109) and will have the same name as my *.Rmd file. 
