---
title: "The Z-test using R"
date: "February 18, 2016"
output: html_document
---

##1. Make up some data; or read in a data file##
```{r}
values = c(61, 78, 66, 57, 79, 81) # make a vector with sample values
```

##2. Look at the sample mean of your variable##
```{r}


```

##3. Set the known parameters for the population##
```{r}
mu <- 75  # set the test value for the population (population mean)
sigma <- 20 # define the population standard devation
```

##4. The z-test allows you to compare the sample mean to a population mean when you know your population parameters and your data are distributed normally. Let's assume the population is normal in shape with a mu = 75 and sigma = 20.##

4.1. You will need the other values, however.
```{r}
# Get the sample mean and assign it to an object
xbar <- mean(values, na.rm = TRUE)

# use length() to obtain how many values are in your sample. If you have NA values, this number will be incorrect.
n <- length(values) 

# take a look at the formula components
mu; sigma; xbar; n
```

4.2. Using those pieces of information, you an calculate by hand or write your own formula for the z-test and assign the result to an object, say zvalue.
```{r}
zvalue <- (xbar - mu)/(sigma/sqrt(n))

# take a look at the value
zvalue

# we can determine the likelihood of the sample mean deviating from the population mean by that many standard errors by getting the p value. 
pvalue <- pnorm(zvalue, mean = 0, sd = 1)

# However, the p-value is one for one tail of the distribution. You must double it. 
pvalue <- pvalue*2
pvalue
```

##5. However, the asbio package will do all these steps for you in one step. Install the asbio package and install the library.##
5.1. The one.sample.z() function will allow you to test this question. You can submit: **(a)** your sample data (values), 
**(b)** your test value under the null hypothesis (mu), **(c)** your population standard deviation (sigma), **(d)** and the sample size (n). By default, the test is two-tailed, but you could change this if you wanted. 

5.2. 
```{r}
library(asbio)
zval2 <- one.sample.z(data = values, null.mu = mu, sigma = sigma, n = n)

# take a look at the z-value and the p-value.
zval2
```

##6. Apply the logic!##
6.1. If population mean = 82, what would your z-test show? 
```{r}
```

6.2. If the population standard deviation = 15, what would your z-test show?
```{r}
```
