---
title: "z_test_population"
author: "Lukas"
date: "Thursday, September 22, 2016"
output: html_document
---


##Part B##
##1. Comparing a sample mean to a population mean##
1.1. The z-test is used to compare a sample mean to a population mean in order to determine if the population from which the sample was drawn is different from another population. The z-test is also only used for variables that are distributed normally. The NHST perspective assumes equivalence between the two populations (e.g., H0: mu1 = mu2), so a difference between the two groups should be 0 on average IF the populations have equivalent means. If, however, the sample produces a mean that differs from the population mean, you need to determine how likely such a result would occur under H0. If the likelihood is below your level of acceptable risk of making a Type I error, you could claim that the populations likely differ. 

1.2. In order to make such a comparision for the z-test, you will need to know the **(a)** population mean, mu, **(b)** the population standard deviation, sigma, **(c)** the sample mean, xbar, and **(d)** the sample size, n. Because the standard deviation of sampling distributions is called the standard error of the mean, the z-test tells you how many standard errors the sampe mean deviates from the population mean. As with extreme z-scores, large z-test values are less likely to occur due to chance if there is really no difference between the population and your sample on your dependent variable. 

1.3. Load the asbio library so that you can use the one.sample.z() function that is part of that library.  
**ANSWER:**
```{r}
#hide
library("asbio")
```

1.4. Read the cdc.csv file the you have used before. Assign it to an object named health.  
**ANSWER:**
```{r}
#hide data <- read.csv("cdc.csv")
health <- read.csv("C:/users/gcook/Desktop/Psyc109/cdc.csv")
```

1.5. If we wanted to test the average heights of people in our sample to that of an average height of a population, we need to specify those values. Obtain the mean height (inches) from your sample data from the cdc.csv file. You can use the mean() function or the summary() function from the psych library. If you use summary(), you will need to load the psych library if you have not already done so. 
**ANSWER:**
```{r}
#hide 
library("psych")
#hide 
mean(health$height)
#hide 
summary(health$height)
```

1.6. You would normally examine whether the heights are distributed normally. We won't do that here, however, doing so is important. Now that you know the sample's mean height, you will need to compare that to the population mean. let's assume that the average adult height of the population is 67 inches tall (5'7") with a standard deviation of 3 inches. Create an object called popmean and assign it a value of 67; create an object called popsd and assign it a value of 3.
**ANSWER:**
```{r}
#hide
popmean <- 67
#hide
popsd <- 3
```

1.7. The cdc data set contains data from very many people (20K). Let' still just assume that's a sample for this exercise. However, your samples will be much smaller of course. In order to calculate the standard error of the mean for the z-test, you will need to know the size of your sample. Use the length() function to obtain the number of people in your height variable and assign that to a variable named samplesize.
**ANSWER:**
```{r}
#hide 
samplesize <- length(health$height)

#take a look at it
samplesize
```

1.8. Now use all of those pieces of information to plug into the one.sample.z() function below where there is a ??. For greatest flexibility, replace the ?? with the objects you created rather than the actual values. The data argument in the function stands for the sample variable you are testing, so I've started it for you by identifying the health data frame. You can copy your modified code and then put it in the R code block to execute. Remember all R code needs to be in those code blocks. 

one.sample.z(data = health$??, null.mu = ??, sigma = ??, n = ??)

**ANSWER:**
```{r}
#hide one.sample.z(data = health$height, null.mu = popmean, sigma = popsd, n = samplesize)
```

1.9. Notice the z-test value and the p-value for the test. Compare your *p*-value to your alpha level and decided whether you would retain or reject H0. State your answer in words.
**ANSWER:** I would reject H0 because the *p*-value is less than an alpha of .05. The difference between the means that that many standard errors is unlikely due to change for a sample of 20K. 



1.10. One thing you will have noticed about this test is that when you have very, very, large sample sizes, the standard error of the mean (sigma/sqrt(n)) will be very, very small. Consequently, you may reject H0 even when the sample mean is very close to the population mean. I used this example to illustrate that point and make you become vigilant about data using very large samples. 
