---
title: 'GCDS Homework #01'
author: ''
date: "Due: 09-08-2022"
output:
  word_document:
    toc: yes
    number_sections: yes
#  html_document:
#    toc: yes
#    number_sections: yes
#    code_folding: hide
#  pdf_document:
#    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

For this homework, you will create an `R Markdown` document file that will output as an HTML file. Name it **"GCDS_HW_01_yourname"** and add your name to it. The homework will need to be submitted as a `knit` HTML file.

```{r eval=FALSE, include=FALSE}
output: 
  html_document:               
    toc: yes                   # include a table of contents
    number_sections: yes       # enumerate sections flagged with #
    code_folding: hide         # allow option to show/hide code
```



You should create a separate code block in `R Markdown` for each of the problems. Type text responses as text and code within code blocks/chunks.




**Questions:**




# Load the `readr` and `ggplot2` libraries.
 
 
```{r include=FALSE}
library(ggplot2)
library(readr)
```
 



# What is the main functionality of `readr`? 




# How to read a file

`read_csv()` requires at least one argument. What is that argument? Is the type of that object you pass to that argument a numeric object or a string/character object? 




# Load a local data frame

Load the `pressure` data set that is part of `base R` (hint: we did this with `cars` and `mtcars` data). 




# Examine a data frame 

Use the `head` function to look at the first several cases/observations of the `pressure` data. 




# Create a plot

Determine what the variable names/columns are and plot any two variables as a scatterplot. You do not know how to plot a scatterplot, so I'll give you some code to help you. You should remember from previous courses that scatter plots plot the x,y coordinates for data points.

The following code uses the `ggplot2` library. If you run this code as is, you will get an error because no arguments are specified. Fix the code so it executes.

`ggplot2::ggplot(pressure, aes(x = ?, y = ?))` `+`
  `geom_point()`


```{r eval=FALSE, include=FALSE}
ggplot2::ggplot(pressure, ggplot2::aes(x = pressure, y = temperature)) +
  ggplot2::geom_point()
    
# or     
library(ggplot2)

ggplot2::ggplot(pressure, aes(x = pressure, y = temperature)) +
  geom_point()
```



# Modify code

Take a look at the following code and the plot it produces. Modify the code below to plot green circles for the xy points. 

Also, make a second plot that swaps the x and y variables, changes the size of the points, and plots a square instead of a circle for each data point. 

```{r}
ggplot(cars, aes(x = speed, y = dist)) +
  geom_point(shape = "circle", 
             color = "darkred", 
             size = 2
             )
```

```{r eval=FALSE, include=FALSE}
ggplot(cars, aes(x = dist, y = speed)) +
  geom_point(shape = "square", 
             color = "green", 
             size = 3
             )
```


# Write data to disk

Use `write_csv()` to write a data frame with file extension `.csv` (to your to your local "GCDS/data" directory on your computer). You will see that we can use `data.frame()` to create a silly data frame, which you can save. When saving data files, remember that you need to specify both the *data frame* object and the *name of the file* as arguments. Importantly, the file name you chose will also need to include the *path* to the subdirectory in which you need to save the file. In other words, you'll need **directorypath/filename + file extension** 

If you query `R` using ?readr::write_csv you'll see an example at the bottom on the help page. Note, however, that there is no subdirectory in that function call. 

`write_csv(mtcars, "mtcars.csv")`

Check also whether you need to pass the data frame object or the file name first. Remember that `R` will by default save to your *working directory*. We set this the other day as "GCDS". You may wish to see whether your working directory includes a "/" at the end of the name by checking what your working directory by checking `getwd()`.  The "/" is used to separate directories.

Data Frame:

```{r}
my_first_data_frame <- data.frame(
  student = c("Bill", "Sally", "Tanya"),   # this is a string vector with 3 elements` 
  age     = c(22, 21, 19)
)
```

If you wanted to write this data frame to your "data" directory as a `.csv` file, what would your code look like? Type it as text below.

```{r eval=FALSE, include=FALSE}
my_first_data_frame <- data.frame(
  student = c("Bill", "Sally", "Tanya"),   # string vector with 3 elements 
  age     = c(22, 21, 19)                  # numeric vector with 3 elements
)

fname = paste0(data_dir,"test.csv" )


write_csv(my_first_data_frame, fname)
```



# Interpreting data

Based on the following plot, what can you tell about the slope?

```{r echo=TRUE}
library(magrittr)   # loading magrittr for using %>%

mtcars %>%
  ggplot(., aes(x = wt, y = mpg)) +
  geom_point()
```




# Thinking of regression

Notice how changing the *aesthetics*, `aes()` in the following code illustrates how cars that have 4, 6, or 8 cylinder engines are represented differently in the plot. Do you believe a single statistical model might fit the data if broken out by cylinder size? Why or why not? Hint: Consider mentally fitting regression lines through the separate points. 

Type your answer as text.



```{r echo=TRUE, message=FALSE, warning=FALSE}
mtcars %>%
  dplyr::mutate(cyl = as.factor(cyl)) %>%  # changing cylinder var to a factor 
  ggplot(., aes(x = wt,                    # x variable
                y = mpg)) +                # y variable
  geom_point(aes(color = cyl))             # make geom_point colors = cyl variable

```

```{r eval=FALSE, include=FALSE}
mtcars %>%
  dplyr::mutate(cyl = as.factor(cyl)) %>%  # changing cylinder var to a factor 
  ggplot(., aes(x = wt,                    # x variable
                y = mpg)) +                # y variable
  geom_point(aes(color = cyl)) +           # make the point colors = cyl variable
  geom_smooth(method = 'lm', 
              se = F,
              fullrange = T) +
  facet_wrap("cyl")                        # create facet for each cylinder
```

**Upload Submission**: Make sure you have named your file correctly, knit it to html (click icon in RStudio) and upload file [here](https://claremontmckenna.app.box.com/f/80863fd8e0f44ec7a95882d9dda8bb91).
