---
title: "Creating Plots"
#author: "Gabriel I. Cook"
#date: "`r Sys.Date()`"
#output: kjhslides::kjh_slides_reader
#editor_options: 
#  chunk_output_type: console
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Creating Plots

## Load our libraries

-   `here` will help keep folders organized

-   [`tidyverse`](https://www.tidyverse.org/packages/) is an ecosystem that will load other relevant libraries

    -   `readr` for reading data files

    -   `dplyr` for manipulating data files

    -   `ggplot` for and creating plots

    -   `magrittr` for piping or chaining code commands


``` R
library(here)
library(tidyverse)
library(magrittr)
```

```{r}
library(here)
library(tidyverse)
library(magrittr)
```


# Getting Data

We will most certainly need data. If you want explore and find some interesting data sets, check out the data archive at the [tidytuesday](https://github.com/rfordatascience/tidytuesday/tree/master/data) github page. They have all data sets corresponding to Tuesday projects dating back to 2018.

# Tidy Data

Speaking of all things tidy, the `ggplot` library can work with data files of different forms but it's particularly designed for tidy data in a long format. You can certainly work with wide-format data but you'll struggle.

## What is *tidy data*?

kjh image img/03_ggplot_tidy_data_only.png

-   A `Variable` is in a *column*
-   A `Case`/data observation is in a *row*
-   A `Cell` contains an observation for a particular variable\
-   Arranged in long format

## Long Format

`Long-format` data frames arrange repeated measurements (e.g., month sales data, daily hours slept, etc.) are on their own rows. For example, the `sales` variable will assume position in a column and the sales corresponding each month will appear on a row. This format results in some data redundancies but each variable assumes only a single column position. Because measurements may occur once or multiple times, they will take positions on rows.

`Wide-format` data frames arrange repeated measurements (e.g., month sales data, daily hours slept, etc.) on separate columns. For example, rather than one variable representing `sales`, each of the 12 month will assume position as its own column (e.g., `sales_jan`, `sales_feb`, etc.).

image would be useful here.

## Untidy data is common for obvious reasons

Here are data presented in wide-format. Notice how rows can contain redundant data. There is a lot of ink used to show nothing new.

```{r }
library(palmerpenguins)
penguins %>% # or also |> 
  dplyr::group_by(species, island, year) %>% 
  dplyr::summarize(bill = round(
    mean(bill_length_mm, na.rm = TRUE), 2)
    ) %>% 
  knitr::kable()
```

Compare that example with the same data in wide-format, which is likely much easier to read. It also take up less real estate; there are fewer places to look. This looking takes effort, time, and can lead to confusion, something we will address with graphical visualizations.

Tabular data are also visualizations. Tables make for efficient comparisons of exact values, which are often difficult to extract from a plot.

-   **example here - side by side?????**

```{r }
penguins %>% # or also |> 
  dplyr::group_by(species, island, year) %>%
  dplyr::summarize(bill = round(mean(bill_length_mm, na.rm = TRUE), 2)) %>%
  tidyr::pivot_wider(names_from = year, values_from = bill) %>%
  knitr::kable()
```

## Problems associated with wide data

-   😠 More than one header row

-   😡 Mixed data types in some columns (extra cleaning)

-   💀 Color and typography used to encode variables and their values

If you want to learn more, check out \[Broman KW, Woo KH (2018) "[Data organization in spreadsheets](doi:10.1080/00031305.2017.1375989)." *The American Statistician* 78:2--10\]

## Rearranging data

If your data are not in long format, you will want to arrange it as such so that `ggplot` can process it easily.

The `tidyr` library will be helpful in such instances.

-   `tidyr::pivot_longer()` for moving from wide to long
-   `tidyr::pivot_wider()` for moving from long to wide

### Rearrange from wide to long using `tidyr::pivot_longer()`

-   Arrange columns `A:C` into a single new column variable, `new_var`
-   The columns will pivot to rows; the file may not be less wide but 3 times as long

```{r pivot-long}
#edu |> 
#  pivot_longer(elem4:coll4, names_to = "education")
```

### Rearrange from wide to long using `tidyr::pivot_longer()`

Decide on a name for the value column

```{r 03-make-a-graph-5 }
#edu |> 
#  pivot_longer(elem4:coll4, 
#               names_to = "education", 
#               values_to = "n")
```
