---
title: 'GCDS Homework #08'
author: ""
date: "Due: 10-20-2022"
output:
  #word_document:
  #  toc: yes
  #  number_sections: yes
  html_document:
    toc: yes
    number_sections: yes
    code_folding: show # vs. hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE, include=FALSE}
########################################################################
# Don't forget to run this
######################################################################
rm(list = ls(all.names = TRUE))      # remove objects in R

source("https://pastebin.com/raw/8mXH02yg")   # run and comment out after knitting
source("https://pastebin.com/raw/97NNTTzu")   # run to include in function definitions

# set the paths for project, script, and data dirs
proj_dir <- gsub("GCDS.*.Rmd", "GCDS", get_ActivePath())
proj_name = ""
r_dir    <- paste(proj_dir, "r", sep = "/")    # r subdir
data_dir <- paste(proj_dir, "data", sep = "/") # data subdir
if ( proj_name != "" & !dir.exists(paste(proj_dir, proj_name, sep = "/")) ) {
  # create project dir
  suppressWarnings(dir.create(paste(proj_dir, proj_name, sep = "/")))
  r_dir <- gsub("/r", paste0("/", proj_name, "/r"), r_dir)
  data_dir <- gsub("/data", paste0("/", proj_name, "/data"), data_dir)
  # create sub directories
  suppressWarnings(dir.create(r_dir))
  suppressWarnings(dir.create(data_dir)) }
```

**Overview**

This homework involves a lot of redundancy for evaluating two models so once you figure out the first model, the second will be very similar.

You will save the `R Markdown` homework document and save it to "GCDS/r" on your computer. Rename as **"GCDS_HW_07_yourname"**. When you open it, add your name to the author part of the YAML code. The homework will need to be submitted as a `knit` HTML file. Feel free to work with a partner but understand that the work submitted has to be your own.


*NOTE: The plots may take a moment to render as there is a lot of data.*


# **Load Libraries**

Load libraries for plots, data manipulation, and model interpretation (e.g., `performance` and `parameters`). If you do not have the `lm.beta` library installed, install it and load it too.

```{r}
library(dplyr)
library(ggplot2)
library(performance)
library(parameters)
```


# **Getting Data**

Take the `diamonds` data from `ggplot2` and select out the variables `x`, `y`, and `z` and assign the data frame to `diamond`.

```{r}
diamond <- diamonds %>% select(., -c(x,y,z)
```



# **Summarize the Data**

What is the mean and price for all diamonds?

```{r}
# grand mean and std
diamond %>%
  summarise(., 
            price = mean(price),
            sd    = sd(price)
            )

```

# **Correlation Matrix**

Select the numeric variables in `diamond` and pass that to `cor()` in order to view the correlation matrix of linear relationships between variables.

```{r}
diamond %>%
  select(., where(is.numeric)) %>%
  cor(.)
```

# **Visualize a Model**

Use `ggplot2` to plot `price` as a function of `carat`. 

```{r}
diamond %>%
  ggplot(., aes(carat, price)) +
  geom_jitter(alpha = .1) + 
  see::theme_modern()
```




# **Simple Linear Regression**

Build a simple linear model that predicts `price` from `carat` size. 

```{r}
lm_carat <- diamond %>%
  lm(price ~ carat, data = .) 
```

# **Check the Model Assumptions**

Use `check_collinearity()`, `check_normality()`, `check_heteroscedasticity()`, `check_autocorrelation()`,  from the `performance` library to check any necessary assumptions. 

Note, if you wanted to see a plot for the checks, wrap the function in `plot()` to obtain the visual.

```{r}
check_normality(lm_carat)
check_heteroscedasticity(lm_carat)
check_autocorrelation(lm_carat)
check_outliers(lm_carat)
```

*Summary*: Describe what do you find.





# **Check the Model Parameters and Performance Metrics**

Using `summary()`, `model_parameters()`, or `model_performance()` to assess the model and then complete the following statements by entering the relevant values.

```{r}
summary(lm_carat)

model_parameters(lm_carat)

model_performance(lm_carat)
```

*Summary*: The percentage of variance in `price` that is explained by variation in `carat` size is _____. Although not realistic, based on the model the mean price for a diamond of `carat` size = 0 would be ____. For each unit increases/decreases in `carat` size, mean `price` of the diamond increases by ___ dollars. 




# **Multiple Predictors**

Now add a diamond's `table` characteristic to the model. You can specify the predictor in a completely new model if that is easy for you.

Alternatively, you can use `update()` to take an existing model and update it with new variable. If the outcome variable is the same, a `.` would take the place of the original outcome variable to the left of the `~`. To keep the first predictor the same, use `.` in place of it after the `~` but then add the other predictors. 

Example: `update(old model, . ~ . + new predictor, data = diamond)`

```{r}
lm_carat_table <- update(lm_carat, . ~ . + table, data = diamond)
```


# **Check the Model Assumptions**

```{r}
check_normality(lm_carat_table)
check_collinearity(lm_carat_table)
check_heteroscedasticity(lm_carat_table)
check_autocorrelation(lm_carat_table)
check_outliers(lm_carat_table)
```


# **Check the Model Parameters and Performance Metrics**

Using `summary()`, `model_parameters()`, or `model_performance()` to assess the model and then complete the following statements by entering the relevant values.

```{r}
summary(lm_carat_table)

model_parameters(lm_carat_table)

model_performance(lm_carat_table)
```

*Summary*: The percentage of variance in `price` that is explained by the predictors is _____. Although not realistic, based on the model the price for a diamond of `carat` size = 0 would be ____. For each unit increase in diamond `carat` size, `price`  increases/decreases by ___ dollars when controlling for (or holding constant) `table` size. Also, for each unit increase in a diamond's `table`, price increases/decreases by ___ dollars when controlling for (or holding constant) `carat` size.


*Comment*: 

Does the estimate associated with `carat` change across the two models? If so, how? 


# **Compare Standardized Coefficients**

Because each predictor has it's own variance in the data, the coefficients can not be directly compared unless they have the same variance, which is unlikely. The standardize slope coefficient can be compared because a change by 1 standard deviation for a variable with a standard deviation of 15 is comparable to 1 standard deviation for a variable with a standard deviation of 72; one standard deviation is one standard deviation.   

Pass the model to `lm.beta::lm.beta()` in order to obtain the standarized coefficients.  

```{r}
lm.beta::lm.beta(lm_carat_table)
```

*Summary*: The percentage of variance in `price` that is explained by the predictors is _____. The price for a diamond of `carat` size = 0 would be ____. For each standard deviation increase in diamond `carat` size, mean `price`  increases/decreases by ___ dollars when controlling for (or holding constant) `table` size. Also, for each standard deviation increase in `table` size, mean price increases/decreases by ___ dollars when controlling for (or holding constant) `carat` size.


Does holding constant the role of `carat` size affect relationship between `table` and `price`? Hint: Look at the correlation matrix? 



# **Compare the Models**

Use `compare_performance()` from the `performance` library to compare your two models.

```{r}
compare_performance(lm_carat, lm_carat_table, rank = T)
```

*Summary*: Does one have a higher R-squared value?; a lower AIC or BIC value?; a lower RMSE value? If so, is one a better model? 




**Upload Submission**: Make sure you have named your file correctly, knit it to html (click icon in RStudio) and upload file [here](https://claremontmckenna.app.box.com/f/80863fd8e0f44ec7a95882d9dda8bb91).

```{r}
message(paste0("knit by ", usr_time))
```


